What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2023-01-01 paper/main.tex
minor change

    \begin{highlights}
        \item We transform Linear Temporal Logic (LTL) formulas to timeline visualizations for the set of satisfying traces.
        \item We translate LTL formulas to \Buchi automata, then to $\omega$-regular expressions, then simplify them and use graph visualization tools to draw timelines.
        \item Our algorithm is sound and complete.
        \item Our tool, \texttt{ltl2timeline}, effectively constructs small timeline visualizations for a range of LTL formulas.
    \end{highlights}

@@

Requirement specification is a central step in the development of safety-critical systems. As a first step, requirements are typically written in natural language. For example, here is a real-world requirement specification from an air traffic control system ~\cite{ZR14}
\begin{center}
    "If a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft"
\end{center}
Such natural language requirements are often ambiguous and not amenable to formal analysis, so requirements must often be specified using rigorous formalized semantics for the purpose of verification and validation. The above requirement, when translated to Linear Temporal Logic (detailed in Section 2.1), may be expressed as follows:
\begin{align*}
    \always (\texttt{tsafe.TSAFE\_command1} \land \texttt{controller.CTR\_control\_1} \\
    \limplies \nextt (\neg \texttt{controller.CTR\_control\_1}))
\end{align*}
Several tools have been developed to reason about formal specifications written with such LTL formulas, like R2U2 and FRET \cite{GPMS20}. However, these formalized semantics are unintintuitive and error-prone for human practitioners, and the tools are much less useful when system designers can't validate input specifications.
Thus, a significant hurdle remains: how can we convincingly demonstrate to the humans in the loop, from system designers to certifiers, that the analyzed formulas truly represent the desired system requirements? % quote MLTL paper?

We take a simple idiomatic approach to address this problem. As Linear Temporal Logic formulas are temporal information, we devise a tool, \texttt{ltl2timeline}, which can draw timelines to depict the satisfying traces of the LTL formula, and thereby represent the specified behaviour in a more natural human-intelligible form.

Our tool, \texttt{ltl2timeline}, uses a remarkably simple algorithm: transforming an input LTL formula through a sequence of automaton and regular expression-based intermediate representations, and is surprisingly effective at providing small representations for a range of LTL formulas. It would visualize the above air traffic control requirement behaviour with the following timeline:
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{examples/ex14/ex14.png}
    \caption{[TODO: make fig with variables, explain a, b or just use oscillating example if that's easier]}
\end{figure}

We make the following contributions in our paper:
\begin{itemize}
    \item We provide the tool, \texttt{ltl2timeline}, which takes LTL formulas and synthesizes timeline visualizations for them
    \item We prove correctness in Section 4. We show that for every LTL formula, the timeline outputted indeed represents the set of satisfying traces.
    \item We showcase the results of our tool on a range of example formulas in Section 5, including discussing the air traffic example above. 
\end{itemize}

@@

\section{Artifact availability}
The tool we present in this paper is available at
\[
    \texttt{\url{https://github.com/EULIR/ltl-explainability}}
\]
which comes with two command-line tools, \texttt{ltl2regex} and \texttt{ltl2timeline}. The specific usage of the tool can be found  \href{https://github.com/EULIR/ltl-explainability#usage}{here}.

@@

\appendix
\section{Concrete Syntax of \texttt{ltl2timeline}} \label{sec:concrete-syntax}
Recall from \ref{ltl-defn} that we define \ltl\ formula as
\begin{align*}
    \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
\end{align*}
where $p \in \AP$. In our tool \texttt{ltl2timeline}, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \Phi & ::=  & p                                     & \texttt{p}                                     & \text{atomic proposition} \\
         & \mid & \neg \Phi                             & \texttt{!} \Phi                                & \text{negation} \\
         & \mid & \Phi \land \Phi                       & \Phi \texttt{ \& } \Phi                        & \text{conjunction} \\
         & \mid & \Phi \lor \Phi                        & \Phi \texttt{ | } \Phi                         & \text{disjunction} \\
         & \mid & \Phi \limplies \Phi                   & \Phi \texttt{ -> } \Phi                        & \text{implication} \\
         & \mid & \always \Phi                          & \texttt{G} \Phi                                & \text{globally} \\
         & \mid & \eventually \Phi                      & \texttt{F} \Phi                                & \text{finally} \\
         & \mid & \nextt \Phi                           & \texttt{X} \Phi                                & \text{next} \\
         & \mid & \Phi \stronguntil \Phi                & \Phi \texttt{ U } \Phi                         & \text{until} \\
         & \mid & \Phi \weakrelease \Phi                & \Phi \texttt{ R } \Phi                         & \text{release} \\
    \end{array}
\]

>>>>

    \begin{highlights}
        \item We transform Linear Temporal Logic (LTL) formulas to timeline visualizations for the set of satisfying traces.
        \item We translate LTL formulas to \Buchi automata, then to $\omega$-regular expressions, then simplify them and use graph visualization tools to draw timelines.
        \item Our algorithm is sound and complete.
        \item Our tool, \tool, effectively constructs small timeline visualizations for a range of LTL formulas.
    \end{highlights}

@@

Requirement specification is a central step in the development of safety-critical systems. As a first step, requirements are typically written in natural language. For example, here is a real-world requirement specification from an air traffic control system ~\cite{ZR14}
\begin{center}
    ``If a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft''
\end{center}
Such natural language requirements are often ambiguous and not amenable to formal analysis, so requirements must often be specified using rigorous formalized semantics for the purpose of verification and validation. The above requirement, when translated to Linear Temporal Logic (detailed in Section 2.1), may be expressed as follows:
\begin{align*}
    \always (\texttt{tsafe.TSAFE\_command1} \land \texttt{controller.CTR\_control\_1} \\
    \limplies \nextt (\neg \texttt{controller.CTR\_control\_1}))
\end{align*}
Several tools have been developed to reason about formal specifications written with such LTL formulas, like R2U2 and FRET \cite{GPMS20}. However, these formalized semantics are unintintuitive and error-prone for human practitioners, and the tools are much less useful when system designers can't validate input specifications.
Thus, a significant hurdle remains: how can we convincingly demonstrate to the humans in the loop, from system designers to certifiers, that the analyzed formulas truly represent the desired system requirements? % quote MLTL paper?

We take a simple idiomatic approach to address this problem. As Linear Temporal Logic formulas are temporal information, we devise a tool, \tool, which can draw timelines to depict the satisfying traces of the LTL formula, and thereby represent the specified behaviour in a more natural human-intelligible form.

Our tool, \tool, uses a remarkably simple algorithm: transforming an input LTL formula through a sequence of automaton and regular expression-based intermediate representations, and is surprisingly effective at providing small representations for a range of LTL formulas. It would visualize the above air traffic control requirement behaviour with the following timeline:
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{examples/ex14/ex14.png}
    \caption{[TODO: make fig with variables, explain a, b or just use oscillating example if that's easier]}
\end{figure}

We make the following contributions in our paper:
\begin{itemize}
    \item We provide the tool, \tool, which takes LTL formulas and synthesizes timeline visualizations for them
    \item We prove correctness in Section 4. We show that for every LTL formula, the timeline outputted indeed represents the set of satisfying traces.
    \item We showcase the results of our tool on a range of example formulas in Section 5, including discussing the air traffic example above.
\end{itemize}

@@

\section{Artifact availability}
The tool we present in this paper is available at
\[
    \texttt{\url{https://github.com/EULIR/ltl-explainability}}
\]
which comes with two command-line tools, \texttt{ltl2regex} and \tool. The specific usage of the tool can be found  \href{https://github.com/EULIR/ltl-explainability#usage}{here}.

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from \ref{ltl-defn} that we define \ltl\ formula as
\begin{align*}
    \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
\end{align*}
where $p \in \AP$. In our tool \tool, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \Phi & ::=  & p                                     & \texttt{p}                                     & \text{atomic proposition} \\
         & \mid & \neg \Phi                             & \texttt{!} \Phi                                & \text{negation} \\
         & \mid & \Phi \land \Phi                       & \Phi \texttt{ \& } \Phi                        & \text{conjunction} \\
         & \mid & \Phi \lor \Phi                        & \Phi \texttt{ | } \Phi                         & \text{disjunction} \\
         & \mid & \Phi \limplies \Phi                   & \Phi \texttt{ -> } \Phi                        & \text{implication} \\
         & \mid & \always \Phi                          & \texttt{G} \Phi                                & \text{globally} \\
         & \mid & \eventually \Phi                      & \texttt{F} \Phi                                & \text{finally} \\
         & \mid & \nextt \Phi                           & \texttt{X} \Phi                                & \text{next} \\
         & \mid & \Phi \stronguntil \Phi                & \Phi \texttt{ U } \Phi                         & \text{until} \\
         & \mid & \Phi \weakrelease \Phi                & \Phi \texttt{ R } \Phi                         & \text{release} \\
    \end{array}
\]
